---
title: "Getting Started with lme4u"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with lme4u}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(lme4u)
library(lme4)
```

## Welcome to `lme4u`

Hierarchical (or mixed-effects) models are essential tools for analyzing data with grouped or nested structure — think students within schools or patients within hospitals. And when it comes to fitting these models in R, the `lme4` package is the go-to hub. 

While `lme4` offers incredible modeling features, it can also be hard to wrap your head around. Many users — especially those transitioning from `stats::lm()` to `lme4::lmer()` or even `lme4::glmer()` — find themselves overwhelmed by the jump in complexity: random intercepts and slopes, link-scale estimates, nested structures, and more. That’s where `lme4u` comes into play.

Built directly on top of lme4, this package is designed to help you **interpret**, **diagnose**, and **hypothesis-test** your mixed models with clarity and confidence. Whether you're a student learning mixed models for the first time or a researcher working with real-world grouped data, `lme4u` walks you through the process with practical explanations and inspiring tools.


## Example Context: Students in Schools

### The Data

`lme4u` comes with a built-in dataset, `star`, a cleaned subset of the well-known [*Project STAR*](https://search.r-project.org/CRAN/refmans/AER/html/STAR.html) (student-teacher achievement ratio) data. The original dataset follows students across elementary school years in Tennessee, aiming to study the effect of class size on student test performance. The processed version in `lme4u` focuses on student performance in third grade, ensuring a hierarchical structure where students are nested within schools.

The shape of `star` is 4192 * 13, which includes a subset of key variables related to student demographics, prior-year (2nd grade) and current-year (3rd grade) test scores, class assignment, teacher qualifications, and school-level identifiers. All students in this dataset have been controlled as attending the same school in both 2nd and 3rd grades. To view details and data dictionary, run `?star` after loading `lme4u` into your workspace.

### The Model

Under the context of `star`, suppose we're intersted in understanding: 

> *How does class type affect student math performance, after accounting for prior achievements?
And does this relationship vary across schools?*

This type of question is common in education research, where students are grouped within schools, and both individual- and group-level heterogeneity contribute to outcomes. In R using `lme4`, this would be implemented as:

```{r}
model <- lmer(math ~ cltype + math_old + (cltype | school_id), data = star)
```

Of course, there could be other ways of modeling the relation. But we will use this specific model to demonstrate how `lme4u` makes interpretation and diagnostics easier throughout the vignette. 


## Interpreting the Model

Immediately after the model fit, you might start wondering whether the model structure is appropriate, or simply, what the formula means statistically. A core component of the package, `explain_lmer()`, provides structured, user-friendly interpretations of linear mixed effects models fitted using `lme4::lmer()`. This function is designed to bridge the gap between raw model output and intuitive understanding, making it easier for users to interpret their model structure and results.

### General Summary

`explain_lmer()` supports three levels of interpretation via the `details` argument. By default, `"general"` will be run if no argument supplied to `details`, which provides a high-level summary of the model, describing its structure, fixed effects, and random effects in a concise yet informative manner:

```{r}
explain_lmer(model)
```

### Fixed Effects

If `"fixed"` is input, the function offers a detailed explanation of the fixed effects, including interpretations of individual estimates and their implications within the model context:

```{r}
fix <- explain_lmer(model, details = "fixed")
```

```{r}
attr(fix, "fixed_details")
```

**NOTE**: If interaction is fitted to the model, the function is also able to detect and adjust the interpretation.

### Random Effects

If `"random"` is input, the function focuses on random effects, explaining random intercepts, slopes, and variance components to highlight across-group variability:

```{r}
random <- explain_lmer(model, details = "random")
```

```{r}
attr(random, "var_details")
```

**NOTE**: The random part is also capable of detecting nested grouping structure, uncorrelated random effects, and multi-level random structures. However, use the interpretation deliberately under such complex cases.


## Assumption Diagnostics

(TBD)

## Testing Effects

Once you’ve interpreted your model and checked that assumptions reasonably hold, the next natural step is to test whether certain effects — fixed or random — meaningfully contribute to your model. `lrt_lmer()` provides likelihood ratio tests on a selected fixed or random effect in a fitted `lme4::lmer()` model. It aims to gauge the effect brought in by one single term and apply statistically accurate null distribution based on different effect types.

Depending on whether you're testing a fixed effect or a random effect by the `type` argument, the structure of the reduced model differs. Currently, the function only supports lmer models with fixed effects and a **single** random structure (i.e., no nested `/`, double vertical `||`, or multi-level random effects). To see more about the testing logic and how the reduced models are constructed differently, run `?lrt_lmer`. 

### Fixed Effects

To test whether a selected fixed effect exists, specify the exact variable name in `target` and set `type = "fixed"`. This will allow the function to match with the underlying set of fixed terms. Running the function will give a summary of output, along with the test statistics being returned as a list:

```{r}
fix_result <- lrt_lmer(model, target = "math_old", type = "fixed", data = star)
```

```{r}
fix_result
```

**NOTE**: If interaction is fitted to the model, specify `has_interaction = TRUE` in the input argument. The reduced model will be subsequently constructed without the selected fixed effect and any interaction terms containing it. 

### Random Effects

To test the excess heterogeneity contributed by a random effect term, specify the exact variable name in `target` and set `type = "random"`. Then, the function will check whether the target is a random intercept (grouping variable) or a random slope in the model, and apply corresponding computation for the likelihood ratio test.

**_Random Intercept_**: The full model is reconstructed by modifying the original `lme4::lmer()` call to retain only the random intercept (ignoring any random slopes). The reduced model is a standard `stats::lm()` model containing only the fixed effects, and the likelihood ratio test follows a mixture Chi-square distribution for p-value computation.

```{r}
random_result1 <- lrt_lmer(model, target = "school_id", type = "random", data = star)
```

**_Random Slope_**: The target variable is only removed from the random slope structure. The reduced model remains an `lme4::lmer()` model, and the likelihood ratio test follows a mixture Chi-square distribution for p-value computation.

```{r}
random_result2 <- lrt_lmer(model, target = "cltype", type = "random", data = star)
```




















